---
title: "Donations Dashboard"
output:
  flexdashboard::flex_dashboard:
    orientation: row
    vertical_layout: scroll
resource_files:
- styles.css
css: styles.css
---

```{r setup, include=FALSE}

library(plyr)    # for data manipulation
library(ggplot2)  # for awesome graphics
library(plotly)   # for interactive graphics
library(flexdashboard)   # for Dashboards
library(data.table) # primary data type for tabular data
library(kableExtra) # kable for printing tabular data
library(scales)     # print values

# Load data
load(file = "./data/donate.RData")

# Compute Critical Values
# Total number of contributions in 2014
computeContributions = function(...) {
  return(  contributions[year == "2014", N] )
} 

# Value of all 2014 contributions
computeValue = function(...) {
  value <- donations_dt[year == "2014", sum(amount)]
  value <- dollar(value)
  return(value)
}

# number of donors in 2014
computeDonors = function(...) {
  return( donor_by_year[ year_num == 2014, .N] )
}

computeRetention = function (...) {
  donors_13 <- donor_by_year[year_num == 2013, name]
  donors_14 <- donor_by_year[year_num == 2014, name]

  # donors who gave in 2013 and in 2014
  num_retain <- intersect(donors_14, donors_13) %>% 
    length

  retain_2014 <- num_retain/length(donors_14) 
  return(retain_2014)
}

```

# Summary

## Plots

### 1. Data Summary {data-width=600}


```{r}

g <- ggplot(donations_dt[amount <= 10000000], 
       aes(x=year, y=amount)) + 
  geom_boxplot(color="darkblue", fill="blue", 
               alpha=0.2, outlier.size = 1) +
  labs(title = "", subtitle = "majority below $500,000",
       x="Year", 
       y = "Amount") +
  theme(legend.position="none") +
  # amount in Millions
	scale_y_continuous(labels = label_number(suffix = " M", 
	                                         scale = 1e-6)) +  
    geom_hline(yintercept = 500000, linetype= "dashed", 
             color = "red")  +
  coord_flip() 
g
```


### 2. Value of Individual Donations {data-width=400, data-height=200}


```{r}
amount_med <- median(donations_dt$amount)

g3 <- ggplot(donations_dt[amount <= 500000], aes(x = amount)) +
  geom_histogram(bins = 55, color="darkblue", fill="white" )+
    labs(x="Amount", y = "Count") +
    geom_vline(aes(xintercept = amount_med), 
               linetype= "dashed", 
             color = "red") +
    scale_x_continuous(label=comma ) +
  annotate(geom="text", x = 200000, y = 2600, label="$60,000 median value",
              color="red") 

ggplotly(g3)
```

## Details

### Note {data-width=600}
The Data Summary shows individual donations up to a value of $10,000,000.  There are an additional 191 donations beyond that value.  These are not shown since they tend to skew the data.

### Donation Amount {data-width=400}
Chart 1 shows that there are some very high value donations of the order of \$10 M.  However, it is better to have many small donations.  Chart 2 shows that most donations are in the range of \$60,000

# Growth Indicators

## Critical Values

### Total Contributions 2014

```{r}
num_contributions = computeContributions()
valueBox(
  num_contributions, icon="fa-reorder", 
  color = "success")
```


### Value of Contributions 2014

####

```{r}
value = computeValue()
valueBox(
  value, icon = "fa-money"
)
```

### Total Donors 2014

```{r}
donors = computeDonors()
valueBox(donors, icon = "fa-users")
```

### Donor Retention 2013 - 2014

```{r}
retention = computeRetention() %>% 
  scales::percent()
#valueBox(scales::percent(retention), icon = "fa-refresh",
#         color = "danger")

gauge(retention, min = 0, max = 100, symbol = '%', gaugeSectors(
  success = c(80, 100), warning = c(40, 79), danger = c(0, 39)
))
```

## Plots

### Donations Growth

```{r}
# Donations Growth
g2 <- ggplot(donate_year, 
             aes(x=year, y = year_sum)) +
  geom_point(color = "blue") +
  labs(title="Total Amount per Year",
       x="Year", 
       y = "Annual Amount") +
  theme(legend.position="none") +
  # amount in thousands
  # scale_y_continuous(label=comma ) +
  # amount in Millions
	scale_y_continuous(labels = label_number(suffix = " M", 
	                                         scale = 1e-6),
	                   n.breaks=6) +  
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) 

ggplotly(g2)
```


### Trend

```{r trend, message=FALSE}
library(mgcv)
g4 <- ggplot(contributions, aes(x = year, y = N)) +
  geom_point(shape=18, color="blue") +
  labs(title="Trend in Number of Donors",
       x="Year", y = "Count") +
  geom_smooth(method = "gam", se=FALSE, color = "orange",
              formula = y ~ s(x) ) + 
  scale_x_continuous(n.breaks=12) +
  annotate(geom="text", x = 2012, y = 3400, 
           label="High in 2012 \n 3776 Donors",
              color="darkblue") 
ggplotly(g4)
```

